### 2018-02-08 chapters/02-synth.tex
add R code to synth

\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach}
\label{chap:synth}

@@

São Paulo has traditionally occupied a key position in Brazil's violence statistics. It is the country's richest and most-densely populated state, and in the 1990s its homicide rate was roughly 50\% higher than the national average \citep[p. 120]{barata2000}. Some areas of the namesake capital city had even worse numbers. Between 1996 and 1999, the ramshackle districts of Jardim São Luiz and Jardim Ângela had respectively 103 and 116 violent deaths per 100,000 residents \citep[p. 8]{cardia2003}, figures that placed them amongst the deadliest neighbourhoods on the globe \citep{who2015}.

@@

Deterrence measures have been complemented by investments in police intelligence. In 1999, the state administration created a new system for crime prevention, Infocrim \citep[p. 3]{risso2014intentional} The system gathers geo-coded information on homicides and maps the most important `hot spots' of criminal activity in the state. The government has also developed a new photo database, Fotocrim, to speed up the process of facial recognition of criminals \citep[p. 3]{mello2010}.

@@

There is good evidence that incapacitation measures have worked well in São Paulo during the last decade. Gun-related homicides have declined about 74\% from 2001 to 2008 \citep{peres2011queda} at the same time when São Paulo has experienced an increase of 770\% in the arrests of repeated murderers \citep[p. 36]{manso2012homicidos}. Although there are studies that indicate possible `hardening effects' of imprisonment, that is, longer sentences may positively affect an individual's tendency to commit further crimes \citep[e.g.][]{chen2007harsher,glaeser1996crime,western2001labor}, the São Paulo case appears to suggest otherwise. Moreover, violent deaths have decreased in all population strata, but especially amongst males (-74.5\%), 15 to 24 year-old men (-78,0\%) and those who live in extreme poverty (-79,3\%), groups that are generally associated with criminality \citep{peres2011queda}.

@@

A group of scholars argue that the PCC significantly contributed to the reduction in violence mainly through the São Paulo prison system. At least since the mid-2000s, these authors argue that the PCC has been able to emerge as an undisputed mediator and solve conflicts between inmates. \citet[p. 83]{dias2009ocupando} writes that ``[\dots] when unable to constitute a universal source of regulation, the official law leaves gaps which are filled by informal instances -- such as the Primeiro Comando da Capital (PCC), in the prisons of São Paulo.'' The gang has implemented informal courts that resemble state institutions, and those meetings have progressively replaced other forms of popular justice such as lynchings or the hiring of target killers \citep[p. 3]{feltran2012metodos}. Moreover, the \emph{Comando} has developed a series of assertive ways to terrorise inmates. Since the PCC's threats are credible, the group is able to impose discipline within the São Paulo prison system \citep{biondi2010,dias2009}.

@@

The difference between direct and total effects can be understood as follows. The direct effect captures the sensitivity of a dependent variable $Y$ to changes in $X$ when this relationship is not mediated by any other variables in the model. Holding all factors constant, the direct effect is a causal chain of length one \citep[p. 160]{sobel1987direct} and could be described simply as $X \rightarrow Y$. In turn, the total effect can be defined as $P(Y_{x} = y)$, that is, ``the probability that response variable $Y$ would take on the value $y$ when $X$ is set to $x$ by external intervention'' \citep[p. 1572]{pearl2001direct}. The total effect is the sum of direct and indirect (or mediated) effects.

@@

The method works as follows.\footnote{Please refer to Appendix~\ref{sec:appendix_a} for a formal presentation of the synthetic control method.} SCM starts with the assumption that one case in the sample has received a treatment. The treatment is defined as a time-delimited event that affects the unit of interest, such as the implementation of a new policy or the outbreak of a conflict. SCM also requires a series of control cases to estimate the models, that is, units that did not receive the treatment during the same period. These cases are often related to the treatment case in some meaningful way, and natural choices for the donor pool are provinces within the same country, or states that share important characteristics. These traits can also be more specifically defined and included as quantitative variables in the estimation models.

@@

Figure \ref{fig:figure7} presents the same test displayed in figure \ref{fig:figure6}, but it uses a stricter threshold for the simulated synthetic controls. The graph features cases in which the mean squared prediction error, a measure of goodness-of-fit, is no higher than twice that of São Paulo. That is, only placebos that have good synthetic matches were selected for the analysis \citep[p. 503]{abadie2010}.  In this group, the negative gap for the homicide rate São Paulo is by far the most relevant, providing further evidence for the original findings.

@@

Lastly, I estimate another synthetic control using another approach. Here, I employ a Bayesian structural time-series model to verify the stability of the previous results \citep{brodersen2015inferring}. This inference procedure is similar to that described in section \ref{sec:methods} and it also consists of matching pre-treatment values of the unit of interest, São Paulo, to other potential control states. However, in this model only the time trends of the dependent variable are matched. In a sense, this is closer to a traditional differences-in-differences approach, but without the restrictive assumption that the treated and the control cases would follow parallel trends over time \citep[p. 494]{abadie2010}.

@@

\section{Appendix} 
\label{sec:appendix_a}

@@

for all $t \in \left\{T_{0} + 1, \dots, T \right\}$, that is, after the intervention period (1999--2009). In practice, $W^{*}$ is chosen non-parametrically as to minimise $||X_{1} - X_{0}W||$, subject to the weight constrains. Consider $||X_{1} - X_{0}W||v = \sqrt{(X_{1} - X_{0}W)'V (X_{1} - X_{0}W)}$, where $V$ is a $(k \times k)$ symmetric and semi-definite positive matrix with the relative importance of each assigned homicide rate predictor. From various possible ways of choosing $V$, in this paper I follow the recommendation of \citet{abadie2003} and choose $V^{*}$ as the value of $V$ that minimises the root mean squared prediction error (RMSPE) for homicide rates in the entire pre-treatment period (1990-1998).

>>>>

\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach\footnote{Forthcoming in \textit{Latin American Research Review}, 53(2), 2018. All data, code, and information required to replicate this study are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.}}
\label{chap:synth}

@@

São Paulo has traditionally occupied a key position in Brazil's violence statistics. It is the country's richest and most-densely populated state, and in the 1990s its homicide rate was roughly 50\% higher than the national average \citep[120]{barata2000}. Some areas of the namesake capital city had even worse numbers. Between 1996 and 1999, the ramshackle districts of Jardim São Luiz and Jardim Ângela had respectively 103 and 116 violent deaths per 100,000 residents \citep[8]{cardia2003}, figures that placed them amongst the deadliest neighbourhoods on the globe \citep{who2015}.

@@

Deterrence measures have been complemented by investments in police intelligence. In 1999, the state administration created a new system for crime prevention, Infocrim \citep[3]{risso2014intentional} The system gathers geo-coded information on homicides and maps the most important `hot spots' of criminal activity in the state. The government has also developed a new photo database, Fotocrim, to speed up the process of facial recognition of criminals \citep[3]{mello2010}.

@@

There is good evidence that incapacitation measures have worked well in São Paulo during the last decade. Gun-related homicides have declined about 74\% from 2001 to 2008 \citep{peres2011queda} at the same time when São Paulo has experienced an increase of 770\% in the arrests of repeated murderers \citep[36]{manso2012homicidos}. Although there are studies that indicate possible `hardening effects' of imprisonment, that is, longer sentences may positively affect an individual's tendency to commit further crimes \citep[e.g.][]{chen2007harsher,glaeser1996crime,western2001labor}, the São Paulo case appears to suggest otherwise. Moreover, violent deaths have decreased in all population strata, but especially amongst males (-74.5\%), 15 to 24 year-old men (-78,0\%) and those who live in extreme poverty (-79,3\%), groups that are generally associated with criminality \citep{peres2011queda}.

@@

A group of scholars argue that the PCC significantly contributed to the reduction in violence mainly through the São Paulo prison system. At least since the mid-2000s, these authors argue that the PCC has been able to emerge as an undisputed mediator and solve conflicts between inmates. \citet[83]{dias2009ocupando} writes that ``[\dots] when unable to constitute a universal source of regulation, the official law leaves gaps which are filled by informal instances -- such as the Primeiro Comando da Capital (PCC), in the prisons of São Paulo.'' The gang has implemented informal courts that resemble state institutions, and those meetings have progressively replaced other forms of popular justice such as lynchings or the hiring of target killers \citep[3]{feltran2012metodos}. Moreover, the \emph{Comando} has developed a series of assertive ways to terrorise inmates. Since the PCC's threats are credible, the group is able to impose discipline within the São Paulo prison system \citep{biondi2010,dias2009}.

@@

The difference between direct and total effects can be understood as follows. The direct effect captures the sensitivity of a dependent variable $Y$ to changes in $X$ when this relationship is not mediated by any other variables in the model. Holding all factors constant, the direct effect is a causal chain of length one \citep[160]{sobel1987direct} and could be described simply as $X \rightarrow Y$. In turn, the total effect can be defined as $P(Y_{x} = y)$, that is, ``the probability that response variable $Y$ would take on the value $y$ when $X$ is set to $x$ by external intervention'' \citep[1572]{pearl2001direct}. The total effect is the sum of direct and indirect (or mediated) effects.

@@

The method works as follows.\footnote{Please refer to Appendix~\ref{sec:synth-appendix} for a formal presentation of the synthetic control method.} SCM starts with the assumption that one case in the sample has received a treatment. The treatment is defined as a time-delimited event that affects the unit of interest, such as the implementation of a new policy or the outbreak of a conflict. SCM also requires a series of control cases to estimate the models, that is, units that did not receive the treatment during the same period. These cases are often related to the treatment case in some meaningful way, and natural choices for the donor pool are provinces within the same country, or states that share important characteristics. These traits can also be more specifically defined and included as quantitative variables in the estimation models.

@@

Figure \ref{fig:figure7} presents the same test displayed in figure \ref{fig:figure6}, but it uses a stricter threshold for the simulated synthetic controls. The graph features cases in which the mean squared prediction error, a measure of goodness-of-fit, is no higher than twice that of São Paulo. That is, only placebos that have good synthetic matches were selected for the analysis \citep[503]{abadie2010}.  In this group, the negative gap for the homicide rate São Paulo is by far the most relevant, providing further evidence for the original findings.

@@

Lastly, I estimate another synthetic control using another approach. Here, I employ a Bayesian structural time-series model to verify the stability of the previous results \citep{brodersen2015inferring}. This inference procedure is similar to that described in section \ref{sec:methods} and it also consists of matching pre-treatment values of the unit of interest, São Paulo, to other potential control states. However, in this model only the time trends of the dependent variable are matched. In a sense, this is closer to a traditional differences-in-differences approach, but without the restrictive assumption that the treated and the control cases would follow parallel trends over time \citep[494]{abadie2010}.

@@

\section{Appendix} 
\label{sec:synth-appendix}

\subsection{The Synthetic Control Estimator}
\label{sec:synth-estimator}

@@

for all $t \in \left\{T_{0} + 1, \dots, T \right\}$, that is, after the intervention period (1999--2009). In practice, $W^{*}$ is chosen non-parametrically as to minimise $||X_{1} - X_{0}W||$, subject to the weight constrains. Consider $||X_{1} - X_{0}W||v = \sqrt{(X_{1} - X_{0}W)'V (X_{1} - X_{0}W)}$, where $V$ is a $(k \times k)$ symmetric and semi-definite positive matrix with the relative importance of each assigned homicide rate predictor. From various possible ways of choosing $V$, here I follow the recommendation of \citet{abadie2003} and choose $V^{*}$ as the value of $V$ that minimises the root mean squared prediction error (RMSPE) for homicide rates in the entire pre-treatment period (1990-1998).

\subsection{\texttt{R} Code}
\label{sec:synth-code}

The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter. The original data files as well as the final dataset are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.

\singlespacing
\small
\begin{verbatim}
######################
### Data Wrangling ###
######################

# Please set your working directory to the data/ folder

# Clear the workspace
rm(list = ls())

# Load necessary packages
library(reshape2)  # data manipulation

# Dependent variable:
dep <- read.csv("homicide-rates.csv", header = TRUE, skip = 1)

dep.molten <- melt(dep,
                   id.vars = c("Sigla",
                               "Código",
                               "Estado")
                    )

colnames(dep.molten) <- c("abbreviation",
                          "code",
                          "state",
                          "year",
                          "homicide.rates")

dep.molten$year <- as.numeric(substring(dep.molten$year, 2))
\end{verbatim}

\begin{verbatim}
# Independent variables
ind1 <- read.csv("state-gdp-capita.csv", header = TRUE, skip = 1)

ind1.molten <- melt(ind1,
                    id.vars = c("Sigla",
                                "Código",
                                "Estado")
                    )

colnames(ind1.molten) <- c("abbreviation",
                           "code",
                           "state",
                           "year",
                           "state.gdp.capita")

ind1.molten$year <- as.numeric(substring(ind1.molten$year, 2))

ind2 <- read.csv("state-gdp-growth-percentage.csv", header = TRUE, skip = 1)

ind2.molten <- melt(ind2,
                    id.vars = c("Sigla",
                                "Código",
                                "Estado")
                    )

colnames(ind2.molten) <- c("abbreviation",
                           "code",
                           "state",
                           "year",
                           "state.gdp.growth.percent")

ind2.molten$year <- as.numeric(substring(ind2.molten$year, 2))

ind3 <- read.csv("gini.csv", header = TRUE, skip = 1)

ind3.molten <- melt(ind3,
                    id.vars = c("Sigla",
                                "Código",
                                "Estado")
                    )

colnames(ind3.molten) <- c("abbreviation",
                           "code",
                           "state",
                           "year",
                           "gini")

ind3.molten$year <- as.numeric(substring(ind3.molten$year, 2))

ind4 <- read.csv("population-projection.csv",
                 header = TRUE,
                 skip   = 1)

ind4.molten <- melt(ind4,
                    id.vars = c("Sigla",
                                "Código",
                                "Estado")
                    )

colnames(ind4.molten) <- c("abbreviation",
                           "code",
                           "state",
                           "year",
                           "population.projection")

ind4.molten$year <- as.numeric(substring(ind4.molten$year, 2))

ind5 <- read.csv("population-extreme-poverty.csv", header = TRUE, skip = 1)

ind5.molten <- melt(ind5,
                    id.vars = c("Sigla",
                                "Código",
                                "Estado")
                    )

colnames(ind5.molten) <- c("abbreviation",
                           "code",
                           "state",
                           "year",
                           "population.extreme.poverty")

ind5.molten$year <- as.numeric(substring(ind5.molten$year, 2))

ind6 <- read.csv("years-schooling.csv", header = TRUE, skip = 1)

ind6.molten <- melt(ind6,
                    id.vars = c("Sigla",
                                "Código",
                                "Estado")
                    )

colnames(ind6.molten) <- c("abbreviation",
                           "code",
                           "state",
                           "year",
                           "years.schooling")

ind6.molten$year <- as.numeric(substring(ind6.molten$year, 2))

# Merges files
data.list <- list(dep.molten,
                  ind1.molten,
                  ind2.molten,
                  ind3.molten,
                  ind4.molten,
                  ind5.molten,
                  ind6.molten)

data1 <- Reduce(function(...) merge(..., all = TRUE), data.list)

# Subset and sort
data2 <- subset(data1, year >= 1990 & year <= 2009)
data2 <- data2[order(data2$state), ]
rownames(data2) <- NULL

# Count missing observations, calculate their percentage
round(sapply(data2, function(x) length(which(is.na(x)))), 2)
round(sapply(data2, function(x) length(which(is.na(x)))/length(x)), 2)

# Linear imputation of missing values.
data2$gini.imp <- approxfun(seq_along(data2$gini), data2$gini)(seq_along(data2$gini))

data2$population.extreme.poverty.imp <- approxfun(seq_along(data2$population.extreme.poverty),
data2$population.extreme.poverty)(seq_along(data2$population.extreme.poverty))

data2$years.schooling.imp <- approxfun(seq_along(data2$years.schooling),
data2$years.schooling)(seq_along(data2$years.schooling))

# Create proportion.extreme.poverty
data2$proportion.extreme.poverty <- data2$population.extreme.poverty.imp / data2$population.projection

# Transform variables to improve interpretation
data2$population.projection.ln <- log(data2$population.projection)

# Save data as df.csv
write.table(data2,
            "df.csv",
            row.names = FALSE,
            col.names = TRUE,
            sep       = ",")

#####################
### Data Analysis ###
#####################

# Load necessary packages
library(dplyr) # data manipulation
library(Synth) # models

# Load data
df <- read.csv("df.csv", header = TRUE)

# Prepare dataset
df$state <- as.character(df$state) # required by dataprep()

# Plot: Homicide rates for Sao Paulo and Brazil (average)
df1 <- df %>%
        mutate(homicide.sp = ifelse(homicide.rates & state == "São Paulo", homicide.rates, NA)) %>%
        select(year, homicide.sp)

df2 <- df %>%
        mutate(homicide.rates1 = ifelse(homicide.rates & state != "São Paulo", homicide.rates, NA)) %>%
        group_by(year) %>%
        summarise(homicide.br = mean(homicide.rates1, na.rm = TRUE))

setEPS()
postscript(file    = "br.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,     # 17.8 cm
           height  = 5.25)  # 13.3 cm

plot(x = df1$year,
     y = df1$homicide.sp,
     type = "l",
     ylim = c(0, 60),
     xlim = c(1990, 2009),
     xlab = "Year",
     ylab = "Homicide Rates",
     cex = 3,
     lwd = 2,
     xaxs = "i",
     yaxs = "i"
)

lines(df2$year,
      df2$homicide.br,
      lty = 2,
      cex = 3,
      lwd = 2)

arrows(1997, 50, 1999, 50,
       col    = "black",
       length = .1)

text(1995, 50,
     "Policy Change",
     cex = .8)

abline(v   = 1999,
       lty = 2)

legend(x = "bottomleft",
       legend = c("São Paulo",
                  "Brazil (average)"),
       lty    = c("solid", "dashed"),
       cex    = .8,
       bg     = "white",
       lwdc(2, 2)
)

invisible(dev.off())

# Prepare data for synth
dataprep.out <-
        dataprep(df,
                 predictors = c("state.gdp.capita",
                                "state.gdp.growth.percent",
                                "population.projection.ln",
                                "years.schooling.imp"
                                ),
                 special.predictors = list(
                         list("homicide.rates", 1990:1998, "mean"),
                         list("proportion.extreme.poverty", 1990:1998, "mean"),
                         list("gini.imp", 1990:1998, "mean")
                         ),
                 predictors.op = "mean",
                 dependent     = "homicide.rates",
                 unit.variable = "code",
                 time.variable = "year",
                 unit.names.variable   = "state",
                 treatment.identifier  = 35,
                 controls.identifier   = c(11:17, 21:27, 31:33, 41:43, 50:53),
                 time.predictors.prior = c(1990:1998),
                 time.optimize.ssr     = c(1990:1998),
                 time.plot             = c(1990:2009)
                 )

# Run synth
synth.out <- synth(dataprep.out)

# Get result tables
print(synth.tables   <- synth.tab(
        dataprep.res = dataprep.out,
        synth.res    = synth.out)
      )

# Plot: Main model
setEPS()
postscript(file    = "trends.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,     # 17.8 cm
           height  = 5.25)  # 13.3 cm

path.plot(synth.res    = synth.out,
          dataprep.res = dataprep.out,
          Ylab         = c("Homicide Rates"),
          Xlab         = c("Year"),
          Legend       = c("São Paulo","Synthetic São Paulo"),
          Legend.position = c("bottomleft")
)

abline(v   = 1999,
       lty = 2)

arrows(1997, 50, 1999, 50,
       col    = "black",
       length = .1)

text(1995, 50,
     "Policy Change",
     cex = .8)

invisible(dev.off())

# Main model: gaps plot
setEPS()
postscript(file    = "gaps.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,
           height  = 5.25)

gaps.plot(synth.res    = synth.out,
          dataprep.res = dataprep.out,
          Ylab         = c("Gap in Homicide Rates"),
          Xlab         = c("Year"),
          Ylim         = c(-30, 30),
          Main         = ""
)

abline(v   = 1999,
       lty = 2)

arrows(1997, 20, 1999, 20,
       col    = "black",
       length = .1)

text(1995, 20,
     "Policy Change",
     cex = .8)

invisible(dev.off())

## Calculating how many lives were saved during the treatment period

# Weights below retrieved form dataprep.out
# State Code  State Weight  State Name        State Abbreviation
# 42          0.274         Santa Catarina    SC
# 53          0.210         Distrito Federal  DF
# 32          0.209         Espirito Santo    ES
# 33          0.169         Rio de Janeiro    RJ
# 14          0.137         Roraima           RR
# 14          0.001         Pernambuco        PB
# 35          treat         Sao Paulo         SP
\end{verbatim}

\begin{verbatim}
# Get years after policy change
df.2 <- df[which(df$year >= 1999),]

# Calculate total number of deaths in SP
num.deaths.sp <- sum( (df.2$homicide.rates[which(df.2$abbreviation == "SP")])/100000 *
    (df.2$population.projection[which(df.2$abbreviation == "SP")]))

# Calculate estimated number of deaths in Synthetic São Paulo
num.deaths.synthetic.sp <- sum( (0.274 * (df.2$homicide.rates[which(df.2$abbreviation == "SC")])/100000 *
(df.2$population.projection[which(df.2$abbreviation == "SP")]))
                                + (0.210 * (df.2$homicide.rates[which(df.2$abbreviation == "DF")])/100000 *
                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
                                + (0.209 * (df.2$homicide.rates[which(df.2$abbreviation == "ES")])/100000 *
                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
                                + (0.169 * (df.2$homicide.rates[which(df.2$abbreviation == "RJ")])/100000 *
                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
                                + (0.137 * (df.2$homicide.rates[which(df.2$abbreviation == "RR")])/100000 *
                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
                                + (0.001 * (df.2$homicide.rates[which(df.2$abbreviation == "PB")])/100000 *
                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
                                )

lives.saved <- num.deaths.synthetic.sp - num.deaths.sp
lives.saved # Between 1999 and 2009

########################
### Robustness Tests ###
########################

# Prepare dataset
df$state <- as.character(df$state) # required by dataprep()

## Placebo Test -- Control ends in 1994
dataprep.out1 <-
        dataprep(df,
                 predictors = c("state.gdp.capita",
                                "state.gdp.growth.percent",
                                "population.projection.ln",
                                "years.schooling.imp"
                 ),
                 special.predictors = list(
                         list("homicide.rates", 1990:1994, "mean"),
                         list("proportion.extreme.poverty", 1990:1994, "mean"),
                         list("gini.imp", 1990:1994, "mean")
                 ),
                 predictors.op = "mean",
                 dependent     = "homicide.rates",
                 unit.variable = "code",
                 time.variable = "year",
                 unit.names.variable   = "state",
                 treatment.identifier  = 35,
                 controls.identifier   = c(11:17, 21:27, 31:33, 41:43, 50:53),
                 time.predictors.prior = c(1990:1994),
                 time.optimize.ssr     = c(1990:1994),
                 time.plot             = c(1990:1998))

# Run synth
synth.out1 <- synth(dataprep.out1)

# Get result tables
print(synth.tables   <- synth.tab(
        dataprep.res = dataprep.out1,
        synth.res    = synth.out1)
      )

# Placebo test: graph
setEPS()
postscript(file    = "placebo.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,
           height  = 5.25)

path.plot(synth.res       = synth.out1,
          dataprep.res    = dataprep.out1,
          Ylab            = c("Homicide Rates"),
          Xlab            = c("Year"),
          Legend          = c("São Paulo","Synthetic São Paulo"),
          Legend.position = c("bottomleft"),
          Ylim            = c(0, 50)
)

abline(v   = 1995,
       lty = 2)

arrows(1994, 40, 1995, 40,
       col    = "black",
       length = .1)

text(1993, 40,
     "Placebo \nPolicy Change",
     cex = .8)

invisible(dev.off())

## Leave-one-out

# Loop over leave one outs
storegaps <- matrix(NA, length(1990:2009), 4)

colnames(storegaps) <- c(14, 33, 42, 53) # RR, RJ, SC, DF
co <- unique(df$code)
co <- co[-25]

for(k in 1:4){

        # Data prep for training model
        omit <- c(14, 33, 42, 53)[k]

        # Prepare data for synth
        dataprep.out2 <-
                dataprep(df,
                         predictors = c("state.gdp.capita",
                                        "state.gdp.growth.percent",
                                        "population.projection.ln",
                                        "years.schooling.imp"
                         ),
                         special.predictors = list(
                                 list("homicide.rates", 1990:1998, "mean"),
                                 list("proportion.extreme.poverty", 1990:1998, "mean"),
                                 list("gini.imp", 1990:1998, "mean")
                         ),
                         predictors.op = "mean",
                         dependent     = "homicide.rates",
                         unit.variable = "code",
                         time.variable = "year",
                         unit.names.variable   = "state",
                         treatment.identifier  = 35,
                         controls.identifier   = co[-which(co==omit)],
                         time.predictors.prior = c(1990:1998),
                         time.optimize.ssr     = c(1990:1998),
                         time.plot             = c(1990:2009)
                )

        # Run synth
        synth.out2 <- synth(dataprep.out2)

        storegaps[,k] <- (dataprep.out2$Y0%*%synth.out2$solution.w)
} # Close loop over leave one outs

# Leave-one-out: graph
setEPS()
postscript(file    = "leave-one-out.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,
           height  = 5.25)

path.plot(synth.res    = synth.out,
          dataprep.res = dataprep.out,
          Ylab         = c("Homicide Rates"),
          Xlab         = c("Year"),
          Legend       = c("São Paulo","Synthetic São Paulo"),
          Legend.position = c("bottomleft"))

abline(v   = 1999,
       lty = 2)

arrows(1997, 50, 1999, 50,
       col    = "black",
       length = .1)

text(1995, 50,
     "Policy Change",
     cex = .8)

for(i in 1:4){
        lines(1990:2009,
              storegaps[,i],
              col = "darkgrey",
              lty = "solid")
}

lines(1990:2009,
      dataprep.out$Y0plot %*% synth.out$solution.w,
      col = "black",
      lty = "dashed",
      lwd = 2)

legend(x = "bottomleft",
       legend = c("São Paulo",
                  "Synthetic São Paulo",
                  "Synthetic São Paulo (leave-one-out)"
       ),
       lty    = c("solid", "dashed", "solid"),
       col    = c("black", "black", "darkgrey"),
       cex    = .8,
       bg     = "white",
       lwdc(2, 2, 1)
)

invisible(dev.off())

## Permutation test
states <- c(11:17, 21:27, 31:33, 35, 41:43, 50:53)

# Prepare data for synth
results <- list()
results_synth <- list()
gaps <- list()

for (i in states) {
    dataprep.out <-
            dataprep(df,
                     predictors = c("state.gdp.capita",
                                    "state.gdp.growth.percent",
                                    "population.projection.ln",
                                    "years.schooling.imp"
                                    ),
                     special.predictors = list(
                             list("homicide.rates", 1990:1998, "mean"),
                             list("proportion.extreme.poverty", 1990:1998, "mean"),
                             list("gini.imp", 1990:1998, "mean")
                             ),
                     predictors.op = "mean",
                     dependent     = "homicide.rates",
                     unit.variable = "code",
                     time.variable = "year",
                     unit.names.variable   = "state",
                     treatment.identifier  = i,
                     controls.identifier   = states[which(states!=i)],
                     time.predictors.prior = c(1990:1998),
                     time.optimize.ssr     = c(1990:1998),
                     time.plot             = c(1990:2009)
                     )
    results[[as.character(i)]] <- dataprep.out
    results_synth[[as.character(i)]] <- synth(results[[as.character(i)]])
    gaps[[as.character(i)]] <- results[[as.character(i)]]$Y1plot - (results[[as.character(i)]]$Y0plot %*%
    results_synth[[as.character(i)]]$solution.w)

}

## Permutation test
setEPS()
postscript(file    = "permutation-gaps2.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,
           height  = 5.25)

plot(1990:2009,
     ylim = c(-30, 30),
     xlim = c(1990,2009),
     ylab = "Gap in Homicide Rates",
     xlab = "Year"
)

for (i in states) {
        lines(1990:2009,
              gaps[[as.character(i)]],
              col = "lightgrey",
              lty = "solid",
              lwd = 2
        )
}

lines(1990:2009,
      gaps[["35"]], # São Paulo
      col = "black",
      lty = "solid",
      lwd = 2
)

abline(v   = 1999,
       lty = 2)
\end{verbatim}

\begin{verbatim}
abline(h   = 0,
       lty = 1,
       lwd = 1)

arrows(1997, 25, 1999, 25,
       col    = "black",
       length = .1)

text(1995, 25,
     "Policy Change",
     cex = .8)

legend(x = "bottomleft",
       legend = c("São Paulo",
                  "Control States"),
       lty  = c("solid", "solid"),
       col  = c("black", "darkgrey"),
       cex  = .8,
       bg  = "white",
       lwdc(2, 2, 1)
)

invisible(dev.off())

# Permutation graph: states with MSPE no higher than 2x São Paulo's
low.mspe <- c(13, 15, 17, 21, 23, 24, 25, 31, 41:43, 53)

setEPS()
postscript(file    = "low-mspe.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,
           height  = 5.25)

plot(1990:2009,
     ylim = c(-30, 30),
     xlim = c(1990,2009),
     ylab = "Gap in Homicide Rates",
     xlab = "Year"
)

for (i in low.mspe) {
lines(1990:2009,
      gaps[[as.character(i)]],
      col = "lightgrey",
      lty = "solid",
      lwd = 2
      )
}

lines(1990:2009,
      gaps[["35"]], # São Paulo
      col = "black",
      lty = "solid",
      lwd = 2
      )

abline(v   = 1999,
       lty = 2)

abline(h   = 0,
       lty = 1,
       lwd = 1)

arrows(1997, 25, 1999, 25,
       col    = "black",
       length = .1)

text(1995, 25,
     "Policy Change",
     cex = .8)

legend(x = "bottomleft",
       legend = c("São Paulo",
                  "Control States (MSPE Less Than Two Times That of São Paulo)"),
       lty    = c("solid", "solid"),
       col    = c("black", "darkgrey"),
       cex    = .8,
       bg     = "white",
       lwdc(2, 2, 1)
)

invisible(dev.off())

## CausalImpact
# Uncomment the lines below to install necessary packages
# install.packages(c("devtools", "dtw"))
# library(devtools)
# install_github("google/CausalImpact")
# install_github("klarsen1/MarketMatching", build_vignettes=TRUE)

# Load packages
library(CausalImpact)
library(MarketMatching)

# Prepare data
df$year2 <- as.Date(paste(df$year, sep = "", "-01-01"))

# Estimate model
mm <- best_matches(data=df,
                   id_variable="code",
                   date_variable="year2",
                   matching_variable="homicide.rates",
                   parallel=TRUE,
                   warping_limit=1, # warping limit=1
                   dtw_emphasis=1, # rely only on dtw for pre-screening
                   matches=5, # request 5 matches
                   start_match_period="1990-01-01",
                   end_match_period="1998-01-01")

# View best matches
subset(mm$BestMatches, code == 35) # SP

# Results
results <- MarketMatching::inference(matched_markets = mm,
                                     test_market = "35",
                                     end_post_period = "2009-01-01")

# Predictions
results$Predictions

# Plot results
results$PlotActualVersusExpected +
        ggtitle("São Paulo versus Synthetic São Paulo") + theme_bw() +
        geom_line(aes(results$PlotActualVersusExpected$data$test_market),colour="#000099")
results$PlotCumulativeEffect

# Graph
setEPS()
postscript(file    = "causal-impact.eps",
           horiz   = FALSE,
           onefile = FALSE,
           width   = 7,     # 17.8 cm
           height  = 5.25)  # 13.3 cm

plot(x = (1990:2009),
     y = as.numeric(results$Predictions$Response),
     type = "l",
     ylim = c(0, 60),
     xlim = c(1990, 2009),
     xlab = "Year",
     ylab = "Homicide Rates",
     cex = 3,
     lwd = 2)

lines(x = (1990:2009),
      y = as.numeric(results$Predictions$Predicted),
      type = "l",
      lty = 2,
      cex = 3,
      lwd = 2)

arrows(1997, 50, 1999, 50,
       col    = "black",
       length = .1)

text(1995, 50,
     "Policy Change",
     cex = .8)

abline(v   = 1999,
       lty = 2)

legend(x = "bottomleft",
       legend = c("São Paulo",
                  "Brazil (average)"),
       lty    = c("solid", "dashed"),
       cex    = .8,
       bg     = "white",
       lwdc(2, 2)
)

invisible(dev.off())
\end{verbatim}

\doublespacing
\normalsize
